Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-01-21 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
clean up hobbs timingx

There are many ways to structure a Gauss-Newton style nonlinear least squares
program code. In organizing and documenting the nearly half-century of programs
in the Nashlib collection associated with @jncnm79, the author realized that this
variety could be an instructive subject for numerical software designers.

@@

<!-- ?? Check the statements about how things are done and redo tests -->
Note that we form $J^t J$ in `nlsr::nlfb()` to provide the $D$ matrix for 
the scaled stabilization, 
which is quite a lot of computational work. The same program, however, 
uses a QR approach to solving the linearized least squares sub-problem.
To avoid the formation of $D$, an experimental package
`nlsralt` was been built to add functions `nlxbx()` and `nlfbx()` that
use ONLY the fixed element (i.e., unit matrix) stabilization. A
preliminary timing (using @microbench13) with the Hobbs unscaled problem 
(@cnm79, Example 12.2) shows
a slight advantage for `nlsralt` in execution timing with an equivalent solution.

@@

A detail in choosing our approach is the mechanism of inclusion of the
Marquardt (or Marquardt-Nash) stabilization. It turns out to be remarkably
easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
has elements that the square roots of  $\lambda (D + \phi 1_n)$. 
Then $K^t K$ is the appropriate $C$ for the Marquardt version
of the Gauss-Newton. Clearly it is equivalent to augment $J$ by **both** $\lambda D$
and $\lambda \phi 1_n$.

Unfortunately, if we decide to use the stabilization involving $D$, then 
we must perform at least a partial computation of the SSCP matrix,
which is relatively costly. For this reason, it may be preferred to use
the unscaled stabilization using $1_n$ only. It may then be important to
provide an explicit (and fixed) parameter scaling.

?? manual scaling -- is it part of nlsr??

@@

One of the issues with creating the SSCP matrix $A$ is that we lose information in finite arithmetic 
(See @jncnm79, Chapter 5). A QR approach to the traditional Gauss-Newton codes is to solve

@@

That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newtx

To move to the Marquardt variant of the Gauss-Newton, we need to augment $J$. Let us call this augmented matrix
$K$. We will also need to augment $r$ by an appropriate number of zeros. 
For the traditional Marquardt method, we
need to use the square roots of the diagonal elements of $J^t J$, that is, an $n$ by $n$ 
diagonal matrix $Z1$ where

$$ Z1_{k,k}  = \sqrt(\lambda) \sqrt(J^t J)_{k,k}  $$

and $r$ is extended by $n$ zeros. For the Nash modification, we add a further $n$ zeros to $r$ and augment
the $K$ matrix with another $n$ by $n$ block $Z2$ which is a diagonal matrix where the diagonal elements are

@@

Unfortunately, there is a need in each major iteration to compute at least the diagonal elements of the
SSCP matrix. It may, in fact, be more efficient to omit the $Z1$ augmentation. Package `nlsralt`, as
mentioned, does this with functions `nlxbx()` (for formula-specified models) and `nlfbx()` (for problems
specified with `res()` and `jac()` functions). This gives the following preliminary example using the 
Hobbs weed problem.

@@

??FIXME??

>>>>

There are many ways to structure a Gauss-Newton style nonlinear least squares
program code. In organizing and documenting the nearly half-century of programs
in the Nashlib collection associated with @jncnm79 and with similar packages
for the R Project for Statistical Computing, the author realized that an overview
of this variety could be instructive for numerical software designers.

@@

<!-- ?? Check the statements about how things are done and redo tests -->
Note that we form $J^t J$ in `nlsr::nlfb()` to provide the $D$ matrix for 
the scaled stabilization, 
which is quite a lot of computational work. The same program, however, 
uses a QR approach to solving the linearized least squares sub-problem.
To avoid the formation of $D$, an experimental package
`nlsralt` was been built to add functions `nlxbx()` and `nlfbx()` that
use ONLY the fixed element (i.e., unit matrix) stabilization. A
preliminary timing (using @microbench13) with the Hobbs unscaled problem 
(@cnm79, Example 12.2), presented later, shows
a slight advantage for `nlsralt` in execution timing, but we caution
that this is but a single example.

@@

A detail in choosing our approach is the mechanism of inclusion of the
Marquardt (or Marquardt-Nash) stabilization. It turns out to be remarkably
easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
has elements that the square roots of  $\lambda (D + \phi 1_n)$. 
Then $K^t K$ is the appropriate $C$ for the Marquardt version
of the Gauss-Newton. Clearly it is equivalent to augment $J$ by **both** $Z1 = \lambda D$
and $Z2 = \lambda \phi 1_n$.

Unfortunately, if we decide to use the stabilization involving $D$, then 
we must perform at least a partial computation of the SSCP matrix,
which is relatively costly. For this reason, it may be preferred to use
the unscaled stabilization using $1_n$ only. It may then be important to
provide an explicit (and fixed) parameter scaling. The function minimization
package `optimx` (@p-optimx) includes such a possibility, but appears to be
missing from nonlinear least squares possibilities in R.

@@

One of the issues with creating the SSCP matrix $A$ is that we lose information in finite arithmetic 
(See @jncnm79, Chapter 5). As noted above, a QR approach to the traditional Gauss-Newton codes is to solve

@@

That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newton equations.

To move to the Marquardt variant of the Gauss-Newton, we need to augment $J$. Also as described above,
augmented matrix $K$ is a row-wise appending $Z1$ and, for the Nash modification $Z2$, to $J$.
We will also need to augment $r$ by an appropriate number of zeros. 
As a reminder, $Z1$ and $Z2$ are given by

$$ Z1_{k,k}  = \sqrt(\lambda) \sqrt(J^t J)_{k,k}  $$
and

@@

We have noted that, at each major iteration, the construction of $Z1$ requires the diagonal elements of the
SSCP matrix. Thus it could be more efficient to omit the $Z1$ augmentation and use an explicit scaling
in the original setup. Package `nlsralt`, as
mentioned, does this with functions `nlxbx()` (for formula-specified models) and `nlfbx()` (for problems
specified with `res()` and `jac()` functions). This gives the following preliminary example using the 
Hobbs weed problem.

@@
